---
title: "Prioritizing GWAS Results and Identifying Risk SNP-Associated Functional Annotation Tree with `**GPATree**' Package"
subtitle:
- Aastha Khatiwada$^1$, Bethany J. Wolf$^1$, Ayse Selen Yilmaz$^2$, Paula S. Ramos$^{1,3}$, Maciej Pietrzak$^2$, Andrew Lawson$^1$, Kelly J. Hunt$^1$, Hang J. Kim$^4$, Dongjun Chung$^{2}$
author:
- $^1$Department of Public Health Sciences, Medical University of South Carolina, 
- Charleston, South Carolina, USA
- $^2$Department of Biomedical Informatics, The Ohio State University, Columbus, 
- Ohio, USA
- $^3$Department of Medicine, Medical University of South Carolina, Charleston, 
- South Carolina, USA
- $^4$Department of Mathematical Sciences, University of Cincinnati, Cincinnati, 
- Ohio, USA
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output:
  pdf_document:
    number_sections: yes
bibliography: references.bib
fontsize: 10pt
geometry: margin=3cm

vignette: >
  %\VignetteIndexEntry{GPATree-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  # tidy = TRUE,
  prompt = TRUE,
  # split=TRUE,
  # tidy = TRUE,
  comment = ""
)
```
# Overview
<!-- \section{Overview} -->

This vignette provides an introduction to the \texttt{GPATree} package. R package \texttt{GPATree} implements GPA-Tree, a novel statistical approach to prioritize genome-wide association studies (GWAS) results while simultaneously identifying the combinations of functional annotations associated with risk-associated genetic variants. GPA-Tree integrates GWAS summary statistics and functional annotation data within a unified framework, by combining a decision tree algorithm (CART)[@breiman1984classification] within the hierarchical model.

The package can be loaded with the command:


```{r setup, include=TRUE, message=FALSE, warning=FALSE}
library(GPATree)
```
```{r,eval=TRUE, include=FALSE, echo = FALSE}
options(digits = 4)
```
This vignette is organized as follows. Sections \ref{fitting} and \ref{shiny} illustrate the recommended \texttt{GPATree-ShinyGPATree} workflow, which provides convenient and interactive genetic data analysis interface. Advanced users might also find Sections \ref{pruning} – \ref{association} useful and these command lines can be used for integrating GPA-Tree as part of the more comprehensive genetic data analysis workflow, for example.

Please feel free to contact Dongjun Chung at \texttt{chung.911@osu.edu} for any questions or suggestions regarding the `\texttt{GPATree}' package.

# Workflow
<!-- \section{Workflow}\label{workflow} -->

In this vignette, we illustrate the GPA-Tree analysis workflow, using the simulated data provided as the  \texttt{GPATreeExampleData} in the \texttt{GPATree} package. In the simulated data, the number of SNPs is set to $M=10,000$ and the number of functional annotations is set to $K=10$. The GWAS association  $p$-values and the binary functional annotation information are stored in \texttt{GPATreeExampleData\$gwasPval} and \texttt{GPATreeExampleData\$annMat}, respectively. The number of rows in \texttt{GPATreeExampleData\$gwasPval} is assumed to be the same as the number of rows in \texttt{GPATreeExampleData\$annMat}, where the $i$-th ($i=1,...,M$) row of \texttt{gwasPval} and \texttt{annMat} correspond to the same SNP.



```{r include=TRUE, message=FALSE, warning=FALSE, split=TRUE}
data(GPATreeExampleData)
dim(GPATreeExampleData$gwasPval)
head(GPATreeExampleData$gwasPval)
dim(GPATreeExampleData$annMat)
head(GPATreeExampleData$annMat)
```
<!-- \subsection{Fitting the GPA-Tree Model}\label{fitting} -->
## Fitting the GPA-Tree Model \label{fitting}

We can fit the GPA-Tree model using the GWAS association $p$-values (\texttt{GPATreeExampleData\$gwasPval}) and functional annotation data (\texttt{GPATreeExampleData\$annMat}) described above, using the code shown below.

```{r GPATree, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE, message=FALSE}
fit.GPATree <- GPATree(gwasPval = GPATreeExampleData$gwasPval,
                       annMat = GPATreeExampleData$annMat,
                       initAlpha = 0.1,
                       cpTry = 0.005)
```
```{r, eval=TRUE, include=FALSE, echo = FALSE}
options(digits = 4)
```
```{r, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
fit.GPATree
```


<!-- \subsection{ShinyGPATree}\label{shiny} -->

## ShinyGPATree \label{shiny}

The following command can be used to initialize the ShinyGPATree app. ShinyGPATree allows for interactive and dynamic investigation of disease-risk-associated SNPs and functional annotation trees using R Shiny.

```{r, eval=FALSE, include=TRUE, message=FALSE, warning=FALSE, tidy=TRUE, comment=""}
ShinyGPATree(fit.GPATree)
```

<!-- \subsection{Prunning GPA-Tree model fit}\label{pruning} -->
## Advanced use \label{advuse}
### Prunning GPA-Tree model fit \label{pruning}

The following command will prune the GPA-Tree model using any cp value between 0 and 1.

```{r, eval=FALSE, include=TRUE, message=FALSE, warning=FALSE}
fit.GPATree.pruned <- prune(fit.GPATree, 0.001)
```

<!-- \subsection{Functional annotation tree}\label{annotation} -->
### Functional annotation tree \label{annotation}
The following command will plot the GPA-Tree functional annotation tree and provide information about the leaves (terminal nodes) in the tree.

```{r, eval=FALSE, include=TRUE, message=FALSE, warning=FALSE}
plot(fit.GPATree)
```

```{r, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
leaf(fit.GPATree)
```

<!-- \subsection{Association mapping}\label{association} -->
### Association mapping\label{association}

Based on the fitted GPA-Tree model, we can make statistical inference about SNPs by identifying: $(1)$ risk-associated SNPs and (2) the leaves of the GPA-Tree model in which the risk-associated SNP are located in using the code below.
```{r, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
assoc.SNP.GPATree <- assoc(fit.GPATree, 
                           FDR = 0.05, 
                           fdrControl="global")
head(assoc.SNP.GPATree)
table(assoc.SNP.GPATree$P1)
table(assoc.SNP.GPATree$leaf)
table(assoc.SNP.GPATree$P1, assoc.SNP.GPATree$leaf)
```

The \texttt{GPATree\_assoc} function returns two columns. The first column contains binary values where 1 indicates that the SNP is associated with the phenotype and 0 indicates otherwise. The second column provides information regarding the leaf in which the SNP is located in the GPA-Tree plot. The \texttt{GPATree\_assoc} allows both local (\texttt{fdrControl="local"}) and global FDR controls (\texttt{fdrControl="global"}) and users can set the threshold using the argument `\texttt{FDR}'. For \texttt{GPATreeExampleData}, GPA-Tree model identified $1458$ risk SNPs at the nominal global FDR level of $0.05$.


# References

